A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-12 Manuscript/Manuscript-7-MLC.tex
version 8 started

In addition to functioning as a parameter that is tuned for appropriate
analysis, \(\lambda\) can function as a descriptive statistic
characterizing the relative strength of phylogenetic signal in
phenotypic traits to determine the extent to which shared evolutionary
history has influenced trait covariation among taxa. The appeal of
\(\lambda\) as a descriptive statistic is that it serves as the basis
for interpreting ``weak'' versus ``strong'' phylogenetic signal; i.e.,
small versus large values of \(\lambda\), respectively, in a comparative
sense (e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al.
2019). Indeed, statements regarding the strength of phylogenetic signal
based on \(\lambda\) are rather common in the evolutionary literature.
For instance, of the 204 papers published in 2019 that estimated and
reported Pagel's \(\lambda\) (found from a literature survey we
conducted in Google.scholar), 40\% explicitly interpreted the strength
of phylogenetic signal for at least one phenotypic trait. Further,
because nearly half of the 1,572 \(\lambda\) values reported were near 0
or 1 (Figure 1) where the biological interpretation of \(\lambda\) is
known (\(\lambda=0\) represents no phylogenetic signal, while
\(\lambda=1\) is phylogenetic signal as expected under Brownian motion),
this percentage is even higher. \hfill\break

@@

Various other approaches use \(\lambda\) as a parameter that can be
adjusted for inferences that are akin to sensitivity analysis. For
instance, some have performed likelihood ratio tests that compare
observed model fits to those obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019) or evaluated whether observed \(\lambda\) differs from an expected
\(\lambda\), based on confidence intervals generated for the expected
value (Vandelook et al. 2019). Qualitative comparisons of \(\lambda\)
estimates have also been performed for multiple traits on the same
phylogenetic tree to infer whether the strength of phylogenetic signal
is greater in one trait as compared to another (e.g., Liu et al. 2019;
Bai et al. 2019). \hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. However, equating values of \(\lambda\) directly
to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., low number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also implies that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we lack a
general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunately, not all model parameters and descriptive statistics are
effect sizes, and thus many summary measures must first be converted to
statistics with standardized units (i.e., conversion to an effect size)
for meaningful comparison (see Rosenthal 1994). As a consequence, it
follows that only if \(\lambda\) is a statistical effect size can
comparisons of estimates across datasets be interpretable. However, the
calculation and statisticl behavior of \(\lambda\) as an effect size has
not yet been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(\kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(\kappa\)
(\(Z_\kappa\)), implying that \(Z_\kappa\) is a more robust effect size
measure. We also propose a two-sample test statistic that may be used to
compare the strength of phylogenetic signal among datasets, and provide
an empirical example to demonstrate its use. We conclude that estimates
of phylogenetic signal using Pagel's \(\lambda\) are often inaccurate,
and thus interpreting strength of phylogenetic signal in phenotypic
datasets based on this measure is compromised. By contrast, effect sizes
obtained from \(\kappa\) hold promise for characterizing phylogenetic
signal, and for comparing the strength of phylogenetic signal across
datasets.

@@

\hypertarget{the-precision-of-lambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The Precision of \(\lambda\) is
Variable}}{The Precision of \textbackslash lambda is Variable}}\label{the-precision-of-lambda-is-variable}}

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal reflecting Brownian motion (when
\(\lambda_{in} =1\)). For each dataset we then estimated phylogenetic
signal (\(\lambda_{est}\)), and calculated the variance of \(\lambda\)
(\(\sigma^2_\lambda\)) across datasets at each input level of
phylogenetic signal and level of species richness as an estimate of
precision. We verified that the variance of traits simulated had no
effect on phylogenetic signal estimation. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each rescaled phylogeny under a Brownian motion
model of evolution (for PGLS regression). For phylogenetic ANOVA, random
groups (\(X\)) were obtained by simulating a discrete (binary, 0 or 1)
character on each phylogeny. Next, the dependent variable was simulated
in such a manner as to contain a known relationship with \(X\) plus
random error containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). The association between \(Y\) and \(X\) was
modeled using a range of values: \(\beta=(0.0,0.25, 0.5, 0.75,1.0)\),
and the residual error (\(\epsilon\)) was modeled to contain
phylogenetic signal simulated under a Brownian motion model of evolution
on each rescaled phylogeny:
\(\epsilon=\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\): (see Revell
2010 for a similar simulation design). The fit of the phylogenetic
regression was estimated using maximum likelihood, and parameter
estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were obtained. We then
calculated precision estimates (\(\sigma^2_\lambda\)) at each input
level of phylogenetic signal and level of species richness. We verified
that the amount of residual variance simulated had no effect on
\(\sigma^2_\lambda\) but did influence the precision of coefficients
estimated from the linear model (precision increased with smaller
\(\epsilon\), as expected). \hfill\break

@@

\begin{align}
    Z_\kappa=\frac{log(\kappa_{obs})-\hat\mu_{log(\kappa)}}{\hat\sigma_{log(\kappa)}}
\end{align}

@@

An effect size based on \(\lambda\) could be envisioned, which is found
as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}.
\end{align}

@@

This study is not the first to compare \(\lambda\) and \(\kappa\) for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \(\kappa\) values are consistent with those found in
the simulations performed by Münkemüller et al.~(2012), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \(\kappa\) at
intermediate phylogenetic signal strengths was observed (see Fig. 2,
Münkemüller et al. 2012). Recent work by Molina-Venegas and Rodríguez
(2017) found that \(\kappa\) but not \(\lambda\) tended to inflate the
estimate of phylogenetic signal, leading to moderate type I and type II
biases, if polytomic chronograms were used. Their work more thoroughly
addressed previous observations of inflated \(\kappa\) for incompletely
resolved phylogenetic trees (Davies et al. 2012; Münkemüller et al.
2012). An interesting question is whether an inflated \(\kappa\) value
leads to an inflated \(Z_\kappa\) or does a tendency of a particular
tree to inflate estimates of \(\kappa\) also inflate the values in
random permutations of a test, in which case \(Z_\kappa\) is robust to
polytomies? We repeated the analyses in Figure 4, adjusting trees to
have 50\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (2017), and found results were consistent (see Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \(\kappa\) as a descriptive statistic does not inflate
\(Z_\kappa\) as an effect size. Furthermore, because comparison of
effect sizes in a test is a comparison of locations of observed values
in their sampling distributions, which would shift concomitantly because
of this tendency, the \(Z_{12}\) test statistic in equation 4 appears to
be robust in spite of unresolved trees. \hfill\break

>>>>

In PGLS analysis, \(\lambda\) is a parameter that is tuned via
log-likelihood profiling for its estimation. After obtaining an
optimized \(\lambda\) value, \(\hat{\lambda}\), likelihood ratio tests
can be performed to compare observed model fits using \(\hat{\lambda}\)
to those obtained when \(\lambda=0\) or \(\lambda=1\) to infer whether
phylogenetic signal differs from no signal or a Brownian motion (BM)
model of evolutionary divergence, respectively (Freckleton et al. 2002;
Cooper et al. 2010; Bose et al. 2019). Confidence limits on
\(\hat{\lambda}\) can be also estimated based on percentiles of a
\(\chi^2\) distribution and where these values are mapped onto a
log-likelihood profile (\emph{sensu} Freckleton et al. 2002). With
respect to confidence limits, one can ascertain whether \(\lambda=0\) or
\(\lambda=1\) is contained within an interval of \(\hat{\lambda}\), as
an analog to likelihood ratio tests (see, e.g., Vandelook et al. 2019).
It is, therefore, tempting to regard \(\hat{\lambda}\) as a descriptive
statistic that measures the relative strength of phylogenetic signal,
for inferring the extent to which shared evolutionary history has
influenced trait covariation among taxa, is common. The appeal of
\(\hat{\lambda}\) as a descriptive statistic is that it serves as the
basis for interpreting ``weak'' versus ``strong'' phylogenetic signal;
i.e., small versus large values of \(\hat{\lambda}\), respectively, in a
comparative sense (e.g., De Meester et al. 2019; Pintanel et al. 2019;
Su et al. 2019). For example, qualitative comparisons of
\(\hat{\lambda}\) for multiple traits on the same phylogenetic tree
provide a basis for inferring whether the strength of phylogenetic
signal is greater in one trait as compared to another (e.g., Liu et al.
2019; Bai et al. 2019). Indeed, statements regarding the strength of
phylogenetic signal based on \(\hat{\lambda}\) are rather common in the
evolutionary literature. For instance, of the 204 papers published in
2019 that estimated and reported Pagel's \(\lambda\) (found from a
literature survey we conducted in Google.scholar), 40\% explicitly
interpreted the strength of phylogenetic signal for at least one
phenotypic trait. Further, because nearly half of the 1,572
\(\hat{\lambda}\) values reported were near 0 or 1 (Figure 1) where the
biological interpretation of \(\lambda\) is known (\(\lambda=0\)
represents no phylogenetic signal, while \(\lambda=1\) is phylogenetic
signal as expected under BM), this percentage is even higher.
\hfill\break

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\hat{\lambda}\), as \(\lambda\) is a parameter
on a bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. However, equating values of \(\hat{\lambda}\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\hat{\lambda}\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. (In order to be precise, the
log-likelihood profile would have to be peaked rather than flat.)
Therefore, understanding the precision in estimating \(\hat{\lambda}\)
is paramount. One study (Boettiger et al. 2012) found that estimates of
Pagel's \(\lambda\) displayed less variation (i.e., greater precision)
when data were simulated on a large phylogeny (\(N=281\)) as compared to
a small one (\(N=13\)). From this observation it was concluded that
insufficient data (i.e., low number of species) was the underlying cause
of the increased variation across parameter estimates (Boettiger et al.
2012). Indeed, such a pattern is common with statistical estimators, as
summary statistics and parameters are often more precise at greater
sample sizes (Cohen 1988). However, this conclusion could imply that the
shape of log-likelihood profiles are assumed consistent with species
number. Such an assumption and the importance of tree shape have to date
not been verified. Thus, despite widespread use of Pagel's (1999)
\(\lambda\) in macroevolutionary studies, at present, we lack a general
understanding of the precision with which \(\hat{\lambda}\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values, \(\hat{\lambda}\), of this parameter correspond to the actual
strength of the underlying input signal in the data. For this to be the
case, \(\hat{\lambda}\) must be a statistical effect size. Effect sizes
are a measure of the magnitude of a statistical effect in data,
represented on a common scale (Glass 1976; Cohen 1988). Effect sizes
have widespread use in many areas of the quantiative sciences, as they
represent measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunately, not all model parameters and descriptive statistics are
effect sizes, directly, and thus many summary measures must first be
converted to statistics with standardized units (i.e., conversion to an
effect size) for meaningful comparison (see Rosenthal 1994). As a
consequence, it follows that only if \(\hat{\lambda}\) is a statistical
effect size can comparisons of estimates across datasets be
interpretable. However, the calculation and statistical behavior of
\(\hat{\lambda}\) as an effect size has not yet been explored.
\hfill\break

As a prortional random variable bounded by 0 and 1, we might expect that
\(\hat{\lambda}\) follows A Bernoulli distribution; i.e., branch lengths
in a tree are scaled proportionally to the probability that data arise
from a BM process. As such, we would also expect in simulation
experiment that given a known \(\lambda\) value used to generate random
data on a tree: the mean of an empirical sampling distribution of
\(\hat{\lambda}\) would approximately equal \(\lambda\); the standard
error of \(\hat{\lambda}\) would be largest at intermediate values of
\(\lambda\), and predictable over the range of \(\lambda\) with respect
to treesize; the distribution of \(\hat{\lambda}\) would be symmetric at
intermediate values of \(\lambda\) and more skewed toward values of 0 or
1; and that the distribution of \(\hat{\lambda}\) will be more
platykurtic at intermediate values of \(\lambda\), becoming more
leptokurtic toward 0 and 1. These properties seemed to be
superficiallyreasonable from the simulations perfomed by Münkemüller et
al.(2012, see Fig. 2). Their study simulated strength of phylogenetic
signal as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\), rather than from prescribed values of
\(\lambda\). Nonetheless, their results revealed that as a random
variable, \(\hat{\lambda}\) resembled a Bernoulli random variable based
superficially on statistical moments (mean, variance, skewness, and
kurtosis; see their Fig. 2) with respect to strength of phylogenetic
signal, for a given tree size. However, for comparisons among tree sizes
for a given strength of phylogenetic signal, median values of
\(\hat{\lambda}\) were slightly, positively associated with the number
of species in a tree, especially at weaker strength of signals,
suggesting \(\hat{\lambda}\) might not be an accurate statistic across a
range of tree sizes. In order for \(\hat{\lambda}\) to be considered a
reliable statistic as a measure of phylogenetic signal, as a minimum, it
should be accurate and precise, properties that could be measured from
simulation experiments for known values of \(\lambda\).

In this study, we evaluate the precision of \(\hat{\lambda}\) as an
optimized parameter, used as a statistic, for estimating known levels of
phylogenetic signal in phenotypic data (data simulated with known
\(\lambda\)). We use computer simulations with differing numbers of
species, differently shaped phylogenies, and differing input levels of
\(\lambda\), to explore the degree to which \(\hat{\lambda}\) correctly
identifies known levels of phylogenetic signal, \(\lambda\), and under
which circumstances. We find that estimates of \(\hat{\lambda}\) vary
widely for a given input value of \(\lambda\), and that the precision of
\(\hat{\lambda}\) is not constant across its range. Rather, there is
decreased precision when input levels of phylogenetic signal are of
intermediate strength. Additionally, the same values of
\(\hat{\lambda}\) may be obtained from datasets containing vastly
different input levels of \(\lambda\). Thus, \(\hat{\lambda}\) is not a
reliable indicator of the strength of phylogenetic signal in phenotypic
data. We then describe a standardized effect size for measuring the
strength of phylogenetic signal in phenotypic datasets and apply the
concept to two common measures of phylogenetic signal: \(\hat{\lambda}\)
and \(\kappa\). Through simulations we find that the precision of effect
sizes based on \(\hat{\lambda}\) (\(Z_{\lambda}\)) are less reliable
than that those based on \(\kappa\) (\(Z_\kappa\)), implying that
\(Z_\kappa\) is a more robust effect size measure. We also propose a
two-sample test statistic that may be used to compare the strength of
phylogenetic signal among datasets, and provide an empirical example to
demonstrate its use. We conclude that estimates of phylogenetic signal
using Pagel's \(\lambda\) (as a descriptive statistic, not as a PGLS
parameter) are often inaccurate, and thus interpreting strength of
phylogenetic signal in phenotypic datasets based on this measure is
compromised. By contrast, effect sizes obtained from \(\kappa\) hold
promise for characterizing phylogenetic signal, and for comparing the
strength of phylogenetic signal across datasets.

@@

\hypertarget{the-precision-of-hatlambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The Precision of \(\hat{\lambda}\) is
Variable}}{The Precision of \textbackslash{}hat\{\textbackslash{}lambda\} is Variable}}\label{the-precision-of-hatlambda-is-variable}}

We conducted a series of computer simulations to evaluate the precision
of \(\hat{\lambda}\) based on varied input of Pagel's \(\lambda\) to
generate data. Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). For each simulation run, we rescaled the
simulated phylogenies by multiplying the internal branches by
\(\lambda_{in}\), using 21 intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a BM model of evolution to obtain
datasets with differing levels of phylogenetic signal, that ranged from
no phylogenetic signal (when \(\lambda_{in} =0\)), to phylogenetic
signal reflecting a BM model of evolution (when \(\lambda_{in} =1\)).
For each dataset we then estimated phylogenetic signal
(\(\hat{\lambda}\)), and calculated its variance
(\(\sigma^2_{\hat{\lambda}}\)) at each input level of phylogenetic
signal and level of species richness as an estimate of precision. Random
data samples were generated from a normal distribution, i.e.,
\(\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\). We verified that the
level of \(\sigma^2\) had no effect on \(\hat{\lambda}\) or
\(\sigma^2_{\hat{\lambda}}\), but the results we present used
\(\sigma^2=1\). \hfill\break

We also evaluated the precision of \(\hat{\lambda}\) when estimated in
PGLS regression and ANOVA (i.e., \(Y\sim{X}\)). For these simulations,
\(Y=\beta{X}+\epsilon\) was used to generate data with varied
phylogenetic signal and variable associations (\emph{sensu} Revell
2010). In the case of PGLS regression, \(X\) was a variable simulated
under a BM model of evolution, constrained by rescaling phylogeny
according to \(\lambda\), and in the case of ANOVA, \(X\) was a balanced
binary variable (values of 0 and 1), randomly ordered. The strength of
variable associations ranged as \(\beta=0.0,0.25, 0.5, 0.75\) and
\(1.0\). Residuals, \(\epsilon\), were simulated under a BM model of
evolution, constrained by rescaling phylogeny according to \(\lambda\).
Simulated independent variables or residuals were generated from a
normal distribution, i.e.,
\(\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\), with the phylogenetic
covariance matrix, \(\mathbf{C}\), implicitly varied for each tree by
the level of \(\lambda\). The fit of the phylogenetic regression was
estimated using maximum likelihood, and parameter estimates
(\(\hat{\lambda}\) and \(\hat{\beta}\)) were obtained. We then
calculated precision estimates (\(\sigma^2_{\hat{\lambda}}\)) at each
input level of phylogenetic signal and level of species richness. We
verified that the level of \(\sigma^2\) had no effect on
\(\sigma^2_{\hat{\lambda}}\) but did influence the precision of
\(\hat{\beta}\) estimated from the linear model (precision increased
with smaller \(\epsilon\), as expected). The results we present used
\(\sigma^2=1\). \hfill\break

@@

\begin{align}
    Z_\kappa=\frac{log(\kappa_{obs})-\hat\mu_{log(\kappa)}}{\hat\sigma_{log(\kappa)}},
\end{align}

@@

An effect size based on \(\lambda\) is a little more challenging
because, as a statistic, it does not have a sampling distribution on
which to measure standard error. Confidence intervals for \(\lambda\)
are not generated from standard error calculations, but rather for the
values of lambda that intersect the log-likehihood profile for
corresponding percentiles of the \(\chi^2\) distribution used to compare
the compare the putative appropriate model to a null model with
\(\lambda = 0\) {[}citation needed, think Boettiger has a paper on
this{]}. The simplest similar effect size is

\begin{align}
   \lvert Z_{\lambda} \rvert = \sqrt{\chi^2_{\hat{\lambda}}}
\end{align}

where, \(\hat{\lambda}\) is the maximized likelihood value of
\(\lambda\) and \(\chi^{2}_{\hat{\lambda}}\) is the likelihood ratio
statistic for the value. Hypothetically, this equation would be better
represented as \(Z_{\lambda} = d \sqrt{\chi^2_{\hat{\lambda}}}\), where
\(d\) is a binary value \((-1,1)\) to indicate a direction based on
whether \(\hat{\lambda}\) is below or above a critical value of
\(\lambda\), for a quantile from a \(\chi^{2}\) distribution at a
probability of 0.5. The critical value would correspond to a value of
\(Z = 0\); i.e., the strength of phylogenetic signal that might be
observed at the 50th percentile in a distribution of random outcomes,
generated from a model where \(\lambda = 0\). However, in some
preliminary simulations, we found that mapping
\(\chi^2_{\hat{\lambda}}\) values this way did not produce effect sizes
that were symmetrical about \(Z = 0\), as the mapping is not linear and,
especially for small trees, the log-likelihood profiles can be rather
flat, making estimation of \(\hat{\lambda}\) rather imprecise (as we
elaborate more below.) We will heneforth use equation 3, recognizing
that large values of \(Z\) would likely be positive, anyway, based on
the reference model for the likelihood ratio test.

@@

This study is not the first to compare \(\lambda\) and \(\kappa\) for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \(\kappa\) values are consistent with those found in
the simulations performed by Münkemüller et al. (2012), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \(\kappa\) at
intermediate phylogenetic signal strengths was observed (see Fig. 2,
Münkemüller et al. 2012). Recent work by Molina-Venegas and Rodríguez
(2017) found that \(\kappa\) but not \(\lambda\) tended to inflate the
estimate of phylogenetic signal, leading to moderate type I and type II
biases, if polytomic chronograms were used. Their work more thoroughly
addressed previous observations of inflated \(\kappa\) for incompletely
resolved phylogenetic trees (Davies et al. 2012; Münkemüller et al.
2012). An interesting question is whether an inflated \(\kappa\) value
leads to an inflated \(Z_\kappa\) or does a tendency of a particular
tree to inflate estimates of \(\kappa\) also inflate the values in
random permutations of a test, in which case \(Z_\kappa\) is robust to
polytomies? We repeated the analyses in Figure 4, adjusting trees to
have 50\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (2017), and found results were consistent (see Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \(\kappa\) as a descriptive statistic does not inflate
\(Z_\kappa\) as an effect size. Furthermore, because comparison of
effect sizes in a test is a comparison of locations of observed values
in their sampling distributions, which would shift concomitantly because
of this tendency, the \(Z_{12}\) test statistic in equation 4 appears to
be robust in spite of unresolved trees. \hfill\break
